Ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2018-01-24 manuscript/usses_water_ms.tex
tiny edits from ARK

%

\begin{abstract}
\textbf{Background.} Precipitation is predicted to become more variable
in the western United States, meaning years of above and below average
precipitation will become more common. Periods of extreme precipitation
are major drivers of interannual variability in ecosystem functioning in
water limited communities, but how ecosystems respond to these extremes
over the long-term may shift with precipitation means and variances.
Long-term changes in ecosystem functional response could reflect
compensatory changes in species composition or species reaching
physiological thresholds at extreme precipitation levels.

@@

\hypertarget{introduction}{%
\section{INTRODUCTION}\label{introduction}}

\new{Water availability is a major driver of annual net primary productivity (ANPP) in grassland ecosystems}
(Huxman et al. 2004; Hsu, Powell, and Adler 2012).
\new{Therefore, projected changes in precipitation regimes, associated with global climate change, will impact grassland ecosystem functioning.}
At any given site, the relationship between ANPP and water availability
(e.g., soil moisture) can be characterized by regressing historical
observations of ANPP on observations of soil moisture. Fitted functional
responses can then be used to infer how ANPP may change under future
precipitation regimes (e.g., Hsu, Powell, and Adler 2012). A problem
with this approach is that it requires extrapolation if future
precipitation falls outside the historical range of variability (Smith
2011; Peters et al. 2012). For example, the soil moisture-ANPP
relationship may be linear within the historical range of interannual
variation, but could saturate at higher levels of soil moisture.
Saturating relationships are actually common (Hsu, Powell, and Adler
2012; Laureano A Gherardi and Sala 2015), perhaps because other
resources, like nitrogen, become more limiting in wet years than dry
years. Failure to accurately estimate the curvature of the soil
moisture-ANPP relationship will lead to over- or underprediction of ANPP
under extreme precipitation (Peters et al. 2012).

Another problem with relying on historical ecosystem functional
responses to predict impacts of altered precipitation regimes is that
these relationships themselves might shift over the long-term. Shifts in
species identities and/or abundances can alter an ecosystem's functional
response to water availability because different species have different
physiological thresholds. Smith, Knapp, and Collins (2009) introduced
the `Hierarchical Response Framework' for understanding the interplay of
community composition and ecosystem functioning in response to long-term
shifts in resources. In the near term, ecosystem functioning such as
ANPP will reflect the physiological responses of individual species to
the manipulated resource level. For example, ANPP may initially decline
under simulated drought because the community consisted of
drought-intolerant species (Hoover, Knapp, and Smith 2014), but
functioning may recover over longer time spans as new species colonize
or resident species reorder in relative abundance. It is also possible
that ecosystem functioning eventually shifts to a new mean state,
reflecting the suite of species in the new community (Knapp, Briggs, and
Smith 2012).

Experimental manipulations of limiting resources, like precipitation,
offer a way to test how ecosystems will respond to resource levels that
fall outside the historical range of variability (Avolio et al. 2015;
Laureano A. Gherardi and Sala 2015; Knapp et al. 2017). Altering the
amount of precipitation over many years should provide insight into the
time scales at which water-limited ecosystems respond to chronic
resource alteration. We propose four alternative scenarios for the
effect of precipitation manipulation on the ecosystem functional
response to soil moisture based on the Hierarchical Response Framework
(Fig. 1). We define `ecosystem functional response' as the relationship
between available soil moisture and ANPP. The four scenarios are based
on possible outcomes at the community (e.g., community composition) and
ecosystem (e.g., soil moisture-ANPP regression) levels.

@@

All four outcomes are possible in any given ecosystem, but the time
scales at which the different scenarios play out likely differ (Smith,
Knapp, and Collins 2009; Wilcox et al. 2016; Knapp et al. 2017). To
determine these time scales, we need to amass information on how quickly
ecosystem functional responses change in different ecosystems. We also
need to understand whether changes at the ecosystem level are driven by
community level changes or individual level responses.

@@

\hypertarget{precipitation-experiment}{%
\subsection{Precipitation Experiment}\label{precipitation-experiment}}

Drought and irrigation treatments were designed to decrease and increase
the amount of ambient precipitation by 50\%. To achieve this, we used a
system of rain-out shelters and automatic irrigation (Gherardi and Sala
2013). The rain-out shelters consisted of transparent acrylic shingles
1-1.5 m above the ground that covered an area of \(2.5\times2\) m. The
shingles intercepted approximately 50\% of incoming rainfall, which was
channeled into 75 liter containers. Captured rainfall was then pumped
out of the containers and sprayed on to the adjacent irrigation plot via
two suspended sprinklers. Pumping was triggered by float switches once
water levels reached about 20 liters. We disconnected the irrigation
pumps each October and reconnected them each April. The rain-out
shelters remained in place throughout the year.

@@

\new{We then modeled these daily deviations ($\Delta x_{g,d,\text{trt.}}$)}
from ambient conditions using a linear mixed effects model with
independent variables for treatment (irrigation or drought), season
(winter, spring, summer, fall), rainfall, and all two-way interactions.
Using the local daily weather station data, we recorded rainy days as
any day with measureable precipitation or the day after such a day and
with average temperatures above 3\(^{\circ}\)C (to exclude days with
snowfall). We fit the model using the \texttt{lme4::lmer()} function
(Bates et al. 2015) in R (R Core Team 2016), with random effects for
plot group and date. We weighted observations by the number of unique
sensors or spot measurements that were taken in each plot on that day.
\new{We then used the model to predict the deviations from ambient conditions produced in the treated plots on each day of the experiment.  
We added these predicted deviations to the average daily ambient soil moisture to generate predictions for daily soil moisture in all of the treated plots: $\bar{x}_{d, \text{trt.}} = \Delta \bar{x}_{d,\text{trt.}} + \bar{x}_{d,\text{ambient}}$, where $\bar{x}_{d,\text{trt.}}$ is the average predicted daily soil moisture in the treated plots and $\bar{x}_{d,\text{ambient}}$ is the daily ambient soil moisture averaged across all control plots.}
We could only predict soil moisture in the treated plots on days for
which we took at least one ambient soil moisture measurement. This
procedure allowed us to predict daily soil moisture conditions for all
plots, even on those days when some of our direct treatment measurements
were missing due to malfunction of the sensors or data loggers. See
Kleinhesselink (2017) for more details on our approach to estimating
soil moisture.

\new{Following the above procedure, we still lacked soil moisture data for March 2012 (observations did not start unitl April) and for a string of days in 2013 during which the soil moisture probes failed to take readings.
To fill in these gaps, we used a version of the SOILWAT soil moisture model}
(Sala, Lauenroth, and Parton 1992)
\new{that has been specifically designed for semiarid shrublands and grasslands}
(Bradford, Schlaepfer, and Lauenroth 2014).
\new{The model was parameterized using generic sagebrush steppe parameters and local soil texture, soil bulk density, and precipitation data}
(Kleinhesselink 2017).
\new{We used the locally parameterized SOILWAT model to generate daily soil moisture predictions for the duration of the experiment, but only used SOILWAT predictions where there were gaps in our data (Fig. 2B and Appendix 2).
SOILWAT predicted daily soil moisture under ambient conditions similar to our control plots.
We applied the same statistical model and procedure described above to estimate soil moisture in drought and irrigation plots based on control plot conditions.}

\hypertarget{data-collection}{%
\subsection{Data Collection}\label{data-collection}}

We estimated aboveground net primary productivity (ANPP) using a
radiometer to relate ground reflectance to plant biomass (Byrne et al.
2011). We recorded ground reflectance at four wavelengths, two
associated with red reflectance (626 nm and 652 nm) and two associated
with near-infrared reflectance (875 nm and 859 nm). At each plot in each
year, we took four readings of ground reflectances at the above
wavelengths. We also took readings in 12 (2015), 15 (2012, 2013, 2014),
or 16 (2016) calibration plots adjacent to the experimental site, in
which we harvested all aboveground biomass produced in the current year
(we excluded litter and standing dead material), dried it to a constant
weight at 60\(^{\circ}\)C, and weighed it to estimate ANPP. We made
radiometer measurements and harvested at peak green biomass each year,
typically in late June.

@@

Species composition data came from two sources: yearly census maps for
each plot made using a pantograph (Hill 1920) and yearly counts of
annual species in each plot. From these sources, we determined the
density of all annuals and perennials forbs, the basal cover of
perennial grasses, and the canopy cover of shrubs. We made a large
plot-treatment-year by species matrix, where columns were filled with
either cover or density, depending on the measurement made for the
particular species. We standardized the values in each column so we
could directly compare species quantified with different metrics
(density, basal cover, and canopy cover). This puts all abundance values
on the same scale, meaning that common and rare species are weighted
equally. Assuming that rare species will respond to treatments more than
common ones, our approach is biased towards detecting compositional
changes.
\new{We limited our analysis of community data to observations from the permanent exclosure containing our drought and irrigation treatments (see \textbf{Community composition over time}).
This exclosure included six of the control plots and all of the treatment plots, for a total of 22 plots.
The other eight control plots are in other pastures. 
Including these plots would add spatial variation in composition, complicating our goal of describing temporal trends in composition.}

\hypertarget{data-analysis}{%
\subsection{Data Analysis}\label{data-analysis}}

\hypertarget{ecosystem-functional-response}{%
\subsubsection{Ecosystem functional
response}\label{ecosystem-functional-response}}

Our main goal was to test whether the relationship between ANPP and soil
moisture differed among the drought, control, and irrigation treatments.
Based on our own observations and previous work at our study site
(Blaisdell 1958; Dalgleish et al. 2011; Adler, Dalgleish, and Ellner
2012), we chose to use cumulative volumetric water content from March
through June as our metric of soil moisture (hereafter referred to as
`VWC'). We fit a generalized linear mixed effects regression model with
log(ANPP) as the response variable and VWC and treatment as fixed
effects. Plot and year of treatment were included as random effects to
account for non-independence of observations, as described below. We
log-transformed ANPP to reduce heteroscedasticity. Both log(ANPP) and
VWC were standardized to have mean 0 and unit variance before fitting
the model {[}i.e., \((x_i - \bar{x})/\sigma_x\){]}.

@@

\noindent{}These hypotheses are based on evidence that
precipitation-ANPP relationships often saturate with increasing
precipitation (Hsu, Powell, and Adler 2012; Laureano A Gherardi and Sala
2015).

@@

\noindent{}where \(\Sigma\) is the variance-covariance matrix and
\(j(i)\) reads as ``plot \(j\) associated with observation \(i\)''. The
random year effects (\(\boldsymbol{\eta}\)) are drawn from a normal
prior with mean 0 and standard deviation \(\sigma_{\text{year}}\), which
was drawn from a half-Cauchy distribution.
\new{We used a vague prior distribution for each $\beta$: $\boldsymbol{\beta} \sim \text{Normal}(0,5)$.}
A full description of our model and associated R (R Core Team 2016) code
is in Appendix 4.

We fit the model using a Bayesian approach and obtained posterior
estimates of all unknowns via the No-U-Turn Hamiltonian Monte Carlo
sampler in Stan (Stan Development Team 2016b). We used the R package
`rstan' (Stan Development Team 2016a) to link R (R Core Team 2016) to
Stan. We obtained samples from the posterior distribution for all model
parameters from four parallel MCMC chains run for 10,000 iterations,
saving every \(10^{\text{th}}\) sample. Trace plots of all parameters
were visually inspected to ensure well-mixed chains and convergence. We
also made sure all scale reduction factors (\(\hat{R}\) values) were
less than 1.1 (Gelman and Hill 2009).

\hypertarget{sensitivity-of-anpp-over-time}{%
\subsubsection{Sensitivity of ANPP over
time}\label{sensitivity-of-anpp-over-time}}

@@

\new{We define `sensitivity' as $\frac{ \text{ANPP}_{\text{control}} - \text{ANPP}_{\text{treatment}} } { \text{VWC}_{\text{ambient}} - \text{VWC}_{\text{treatment}} }$, following}
Wilcox et al. (2017).
\new{This metric of sensitivity scales the difference between ANPP in the treated and control plots by the change in soil moisture in the treated plots. 
Because our treated plots are not directly paired with individual control plots, we compare ANPP in each treated plot to the average ANPP of the control plots in each year.
After calculating sensitivity for each drought and irrigation plot in each year, we regressed sensitivity against year of treatment using the \texttt{lm()} function in R}
(R Core Team 2016).
\new{This analysis also allows us to link particularly sensitive treatment-years to changes in community composition.}

\hypertarget{community-composition-over-time}{%
\subsubsection{Community composition over
time}\label{community-composition-over-time}}

We used nonmetric multidimensional scaling (NMDS) based on Bray-Curtis
distances to identify temporal changes in community composition among
treatments. We first calculated Bray-Curtis distances among all plots
for each year of the experiment and then extracted those distances for
use in the NMDS. Some values of standardized species' abundances were
negative, which is not allowed for calculating Bray-Curtis distances. We
simply added `2' to each abundance value to ensure all values were
greater than zero. We plotted the first two axes of NMDS scores to see
if community composition overlapped, or not, among treatments in each
year. We used the \texttt{vegan::metaMDS()} function (Oksanen 2016) to
calculate Bray-Curtis distances and then to run the NMDS analysis. We
used the \texttt{vegan::adonis()} function (Oksanen 2016) to perform
permutational multivariate analysis of variance to test whether
treatment plots formed distinct groupings. To test whether treatment
plots were equally dispersed, or not, we used the
\texttt{vegan::betadisper()} function (Oksanen 2016).

@@

\hypertarget{discussion}{%
\section{DISCUSSION}\label{discussion}}

Ecosystem response to a new precipitation regime depends on the
physiological responses of constituent species and the rate at which
community composition shifts to favor species better able to take
advantage of, or cope with, new resource levels (Smith, Knapp, and
Collins 2009). Previous work has shown that community compositional
shifts can be either rapid, on the order of years (Hoover, Knapp, and
Smith 2014), or slow, on the order of decades (Knapp, Briggs, and Smith
2012; Wilcox et al. 2016). A lingering question is how the time scales
of ecosystem response and community change vary among ecosystems.
Precipitation manipulation experiments can help answer this question,
especially if they push water availability outside the historical range
of variability for long periods.

@@

The similarity of ecosystem functional response (Fig. 3) and community
composition (Fig. 4) among treatments is surprising because grasslands
generally, and sagebrush steppe specifically, are considered
water-limited systems. \new{For example,} Huxman et al. (2004) and Knapp
et al. (2015)
\new{showed that semi-arid sites are more sensitive to drought than mesic sites, and}
Wilcox et al. (2017)
\new{found that semi-arid sites are particularly sensitive to irrigation treatments.}
\new{Based on these findings}, we expected ecosystem functional
response, community composition, or both to change under our treatments.
Why did our treatments fail to induce ecosystem or community responses?
We can think of three reasons. Two are limitations of our study, and one
invokes the life history traits of the species at our study site.

First, our precipitation manipulations may not have been large enough to
induce a response. A 50\% decrease in any given year may not be abnormal
given our site's historical range of variability (Knapp et al. 2017). We
cannot definitively rule out this possibility, but we have reason to
believe our treatments \emph{should} have been large enough. Using the
methods described by Lemoine et al. (2016), we calculated the percent
reduction and increase of \new{water-year} precipitation necessary to
reach the 1\% and 99\% extremes of the historical precipitation regime
at our site (Fig. A5-1).
\new{The 1\% quantile of water-year precipitation at our site is 78 mm, a 26\% reduction from the mean, and the 99\% quantile is 545 mm, a 84\% increase from mean growing season precipitation (Appendix 6).
Thus, our drought treatment represented extreme precipitation amounts, and that is the treatment where we observed a small effect on the slope between soil moisture and ANPP (Fig. 3A).
The irrigation treatment may not have been extreme enough to induce a response.}

Second, ANPP at our site is likely influenced by additional factors, not
only the cumulative March-June soil moisture covariate we included in
our statistical model. \new{For example,} La Pierre et al. (2016)
\new{found that site-scale ANPP is better predicted by nutrient availability than precipitation.}
Moreover, temperature can impact ANPP directly (Epstein, Lauenroth, and
Burke 1997) and by exacerbating the effects of soil moisture (De Boeck
et al. 2011). Measurements of soil moisture likely contain a signal of
temperature, through its impact on evaporation and infiltration, but the
measurements will not capture the direct effect of temperature on
metabolic and physiological processes. We also did not redistribute snow
across our treatments in the winter, and snow melt may spur early spring
growth. Failure to account for potentially important covariates could
explain the within-year spread of ANPP (Fig 2C, Fig. 3B) and the
resulting uncertain relationship we observed between soil moisture and
ANPP across all treatments (Table 1, Fig. A3-2).

Third, the life history traits of the dominant species in our study
ecosystem may explain the consistently positive, but weak and uncertain,
effect of soil moisture on ANPP (Table 1, Fig. 3). Species that live in
variable environments, such as cold deserts, must have strategies to
ensure long-term success as conditions vary. One strategy is bet
hedging, where species forego short-term gains to reduce the variance of
long-term success (Seger 1987). In other words, species follow the same
conservative strategy every year, designed to minimize losses during
unfavorable periods. The dry and variable environment of the sagebrush
steppe has likely selected for bet hedging species that can maintain
function at low water availability and have weak responses to high water
availability. In so doing, the dominant species in our ecosystem avoid
``boom and bust'' cycles, which may explain the weak effect of soil
moisture on ANPP (i.e., the Bayesian credible intervals for the slopes
overlapping zero; Table 1).

Another strategy to deal with variable environmental conditions is
avoidance, which would also result in a consistent ecosystem functional
response between drought and control treatments. For example, many of
the perennial grasses in our focal ecosystem avoid drought stress by
growing early in the growing season (Blaisdell 1958, A.R.
Kleinhesselink, personal observation). Furthermore, the dominant shrub
in our focal ecosystem, \emph{Artemisia tripartita}, has access to water
deep in the soil profile thanks to a deep root system (Kulmatiski et al.
2017).
\new{The dominance of our site by the shrub \emph{A. tripartita} (Appendix 1) may explain why our results do not conform to broader patterns of grassland sensitivity to precipitation manipulations}
(e.g., Huxman et al. 2004; Knapp et al. 2015; Wilcox et al. 2017).

@@

\hypertarget{conclusions}{%
\section{CONCLUSIONS}\label{conclusions}}

>>>>

% 
\begin{abstract}
\textbf{Background.} Precipitation is predicted to become more variable
in the western United States, meaning years of above and below average
precipitation will become more common. Periods of extreme precipitation
are major drivers of interannual variability in ecosystem functioning in
water limited communities, but how ecosystems respond to these extremes
over the long-term may shift with precipitation means and variances.
Long-term changes in ecosystem functional response could reflect
compensatory changes in species composition or species reaching
physiological thresholds at extreme precipitation levels.

@@

\section{INTRODUCTION}\label{introduction}

\new{Water availability is a major driver of annual net primary productivity (ANPP) in grassland ecosystems}
\citep{Huxman2004, Hsu2012}.
\new{Therefore, projected changes in precipitation regimes, associated with global climate change, will impact grassland ecosystem functioning.}
At any given site, the relationship between ANPP and water availability
(e.g., soil moisture) can be characterized by regressing historical
observations of ANPP on observations of soil moisture. Fitted functional
responses can then be used to infer how ANPP may change under future
precipitation regimes \citep[e.g.,][]{Hsu2012}. A problem with this
approach is that it requires extrapolation if future precipitation falls
outside the historical range of variability
\citep{Smith2011, Peters2012}. For example, the soil moisture-ANPP
relationship may be linear within the historical range of interannual
variation, but could saturate at higher levels of soil moisture.
Saturating relationships are actually common
\citep{Hsu2012, Gherardi2015a}, perhaps because other resources, like
nitrogen, become more limiting in wet years than dry years. Failure to
accurately estimate the curvature of the soil moisture-ANPP relationship
will lead to over- or underprediction of ANPP under extreme
precipitation \citep{Peters2012}.

Another problem with relying on historical ecosystem functional
responses to predict impacts of altered precipitation regimes is that
these relationships themselves might shift over the long-term. Shifts in
species identities and/or abundances can alter an ecosystem's functional
response to water availability because different species have different
physiological thresholds. \citet{Smith2009} introduced the `Hierarchical
Response Framework' for understanding the interplay of community
composition and ecosystem functioning in response to long-term shifts in
resources. In the near term, ecosystem functioning such as ANPP will
reflect the physiological responses of individual species to the
manipulated resource level. For example, ANPP may initially decline
under simulated drought because the community consisted of
drought-intolerant species \citep{Hoover2014}, but functioning may
recover over longer time spans as new species colonize or resident
species reorder in relative abundance. It is also possible that
ecosystem functioning eventually shifts to a new mean state, reflecting
the suite of species in the new community \citep{Knapp2012}.

Experimental manipulations of limiting resources, like precipitation,
offer a way to test how ecosystems will respond to resource levels that
fall outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. We propose four alternative scenarios for the effect of
precipitation manipulation on the ecosystem functional response to soil
moisture based on the Hierarchical Response Framework (Fig. 1). We
define `ecosystem functional response' as the relationship between
available soil moisture and ANPP. The four scenarios are based on
possible outcomes at the community (e.g., community composition) and
ecosystem (e.g., soil moisture-ANPP regression) levels.

@@

All four outcomes are possible in any given ecosystem, but the time
scales at which the different scenarios play out likely differ
\citep{Smith2009, Wilcox2016, Knapp2017}. To determine these time
scales, we need to amass information on how quickly ecosystem functional
responses change in different ecosystems. We also need to understand
whether changes at the ecosystem level are driven by community level
changes or individual level responses.

@@

\subsection{Precipitation Experiment}\label{precipitation-experiment}

Drought and irrigation treatments were designed to decrease and increase
the amount of ambient precipitation by 50\%. To achieve this, we used a
system of rain-out shelters and automatic irrigation
\citep{Gherardi2013}. The rain-out shelters consisted of transparent
acrylic shingles 1-1.5 m above the ground that covered an area of
\(2.5\times2\) m. The shingles intercepted approximately 50\% of
incoming rainfall, which was channeled into 75 liter containers.
Captured rainfall was then pumped out of the containers and sprayed on
to the adjacent irrigation plot via two suspended sprinklers. Pumping
was triggered by float switches once water levels reached about 20
liters. We disconnected the irrigation pumps each October and
reconnected them each April. The rain-out shelters remained in place
throughout the year.

@@

\new{We then modeled these daily deviations ($\Delta x_{g,d,\text{trt.}}$)}
from ambient conditions using a linear mixed effects model with
independent variables for treatment (irrigation or drought), season
(winter, spring, summer, fall), rainfall, and all two-way interactions.
Using the local daily weather station data, we recorded rainy days as
any day with measureable precipitation or the day after such a day and
with average temperatures above 3\(^{\circ}\)C (to exclude days with
snowfall). We fit the model using the \texttt{lme4::lmer()} function
\citep{Bates2015} in R \citep{R2016}, with random effects for plot group
and date. We weighted observations by the number of unique sensors or
spot measurements that were taken in each plot on that day.
\new{We then used the model to predict the deviations from ambient conditions produced in the treated plots on each day of the experiment.  
We added these predicted deviations to the average daily ambient soil moisture to generate predictions for daily soil moisture in all of the treated plots: $\bar{x}_{d, \text{trt.}} = \Delta \bar{x}_{d,\text{trt.}} + \bar{x}_{d,\text{ambient}}$, where $\bar{x}_{d,\text{trt.}}$ is the average predicted daily soil moisture in the treated plots and $\bar{x}_{d,\text{ambient}}$ is the daily ambient soil moisture averaged across all control plots.}
We could only predict soil moisture in the treated plots on days for
which we took at least one ambient soil moisture measurement. This
procedure allowed us to predict daily soil moisture conditions for all
plots, even on those days when some of our direct treatment measurements
were missing due to malfunction of the sensors or data loggers. See
\citet{Kleinhesselink2017b} for more details on our approach to
estimating soil moisture.

\new{Following the above procedure, we still lacked soil moisture data for March 2012 (observations did not start unitl April) and for a string of days in 2013 during which the soil moisture probes failed to take readings.
To fill in these gaps, we used a version of the SOILWAT soil moisture model}
\citep{Sala1992a}
\new{that has been specifically designed for semiarid shrublands and grasslands}
\citep{Bradford2014}.
\new{The model was parameterized using generic sagebrush steppe parameters and local soil texture, soil bulk density, and precipitation data}
\citep{Kleinhesselink2017b}.
\new{We used the locally parameterized SOILWAT model to generate daily soil moisture predictions for the duration of the experiment, but only used SOILWAT predictions where there were gaps in our data (Fig. 2B and Appendix 2).
SOILWAT predicted daily soil moisture under ambient conditions similar to our control plots.
We applied the same statistical model and procedure described above to estimate soil moisture in drought and irrigation plots based on control plot conditions.}

\subsection{Data Collection}\label{data-collection}

We estimated aboveground net primary productivity (ANPP) using a
radiometer to relate ground reflectance to plant biomass
\citep{Byrne2011}. We recorded ground reflectance at four wavelengths,
two associated with red reflectance (626 nm and 652 nm) and two
associated with near-infrared reflectance (875 nm and 859 nm). At each
plot in each year, we took four readings of ground reflectances at the
above wavelengths. We also took readings in 12 (2015), 15 (2012, 2013,
2014), or 16 (2016) calibration plots adjacent to the experimental site,
in which we harvested all aboveground biomass produced in the current
year (we excluded litter and standing dead material), dried it to a
constant weight at 60\(^{\circ}\)C, and weighed it to estimate ANPP. We
made radiometer measurements and harvested at peak green biomass each
year, typically in late June.

@@

Species composition data came from two sources: yearly census maps for
each plot made using a pantograph \citep{Hill1920} and yearly counts of
annual species in each plot. From these sources, we determined the
density of all annuals and perennials forbs, the basal cover of
perennial grasses, and the canopy cover of shrubs. We made a large
plot-treatment-year by species matrix, where columns were filled with
either cover or density, depending on the measurement made for the
particular species. We standardized the values in each column so we
could directly compare species quantified with different metrics
(density, basal cover, and canopy cover). This puts all abundance values
on the same scale, meaning that common and rare species are weighted
equally. Assuming that rare species will respond to treatments more than
common ones, our approach is biased towards detecting compositional
changes.
\new{We limited our analysis of community data to observations from the permanent exclosure containing our drought and irrigation treatments (see \textbf{Community composition over time}).
This exclosure included six of the control plots and all of the treatment plots, for a total of 22 plots.
The other eight control plots are in other pastures. 
Including these plots would add spatial variation in composition, complicating our goal of describing temporal trends in composition.}

\subsection{Data Analysis}\label{data-analysis}

\subsubsection{Ecosystem functional
response}\label{ecosystem-functional-response}

Our main goal was to test whether the relationship between ANPP and soil
moisture differed among the drought, control, and irrigation treatments.
Based on our own observations and previous work at our study site
\citep{Blaisdell1958, Dalgleish2011, Adler2012}, we chose to use
cumulative volumetric water content from March through June as our
metric of soil moisture (hereafter referred to as `VWC'). We fit a
generalized linear mixed effects regression model with log(ANPP) as the
response variable and VWC and treatment as fixed effects. Plot and year
of treatment were included as random effects to account for
non-independence of observations, as described below. We log-transformed
ANPP to reduce heteroscedasticity. Both log(ANPP) and VWC were
standardized to have mean 0 and unit variance before fitting the model
{[}i.e., \((x_i - \bar{x})/\sigma_x\){]}.

@@

\noindent{}These hypotheses are based on evidence that
precipitation-ANPP relationships often saturate with increasing
precipitation \citep{Hsu2012, Gherardi2015a}.

@@

\noindent{}where \(\Sigma\) is the variance-covariance matrix and
\(j(i)\) reads as ``plot \(j\) associated with observation \(i\)''. The
random year effects (\(\boldsymbol{\eta}\)) are drawn from a normal
prior with mean 0 and standard deviation \(\sigma_{\text{year}}\), which
was drawn from a half-Cauchy distribution.
\new{We used a vague prior distribution for each $\beta$: $\boldsymbol{\beta} \sim \text{Normal}(0,5)$.}
A full description of our model and associated R \citep{R2016} code is
in Appendix 4.

We fit the model using a Bayesian approach and obtained posterior
estimates of all unknowns via the No-U-Turn Hamiltonian Monte Carlo
sampler in Stan \citep{stan2016}. We used the R package `rstan'
\citep{rstan2016} to link R \citep{R2016} to Stan. We obtained samples
from the posterior distribution for all model parameters from four
parallel MCMC chains run for 10,000 iterations, saving every
\(10^{\text{th}}\) sample. Trace plots of all parameters were visually
inspected to ensure well-mixed chains and convergence. We also made sure
all scale reduction factors (\(\hat{R}\) values) were less than 1.1
\citep{Gelman2009}.

\subsubsection{Sensitivity of ANPP over
time}\label{sensitivity-of-anpp-over-time}

@@

\new{We define `sensitivity' as $\frac{ \text{ANPP}_{\text{control}} - \text{ANPP}_{\text{treatment}} } { \text{VWC}_{\text{ambient}} - \text{VWC}_{\text{treatment}} }$, following}
\citet{Wilcox2017}.
\new{This metric of sensitivity scales the difference between ANPP in the treated and control plots by the change in soil moisture in the treated plots. 
Because our treated plots are not directly paired with individual control plots, we compare ANPP in each treated plot to the average ANPP of the control plots in each year.
After calculating sensitivity for each drought and irrigation plot in each year, we regressed sensitivity against year of treatment using the \texttt{lm()} function in R}
\citep{R2016}.
\new{This analysis also allows us to link particularly sensitive treatment-years to changes in community composition.}

\subsubsection{Community composition over
time}\label{community-composition-over-time}

We used nonmetric multidimensional scaling (NMDS) based on Bray-Curtis
distances to identify temporal changes in community composition among
treatments. We first calculated Bray-Curtis distances among all plots
for each year of the experiment and then extracted those distances for
use in the NMDS. Some values of standardized species' abundances were
negative, which is not allowed for calculating Bray-Curtis distances. We
simply added `2' to each abundance value to ensure all values were
greater than zero. We plotted the first two axes of NMDS scores to see
if community composition overlapped, or not, among treatments in each
year. We used the \texttt{vegan::metaMDS()} function \citep{Oksanen2016}
to calculate Bray-Curtis distances and then to run the NMDS analysis. We
used the \texttt{vegan::adonis()} function \citep{Oksanen2016} to
perform permutational multivariate analysis of variance to test whether
treatment plots formed distinct groupings. To test whether treatment
plots were equally dispersed, or not, we used the
\texttt{vegan::betadisper()} function \citep{Oksanen2016}.

@@

\section{DISCUSSION}\label{discussion}

Ecosystem response to a new precipitation regime depends on the
physiological responses of constituent species and the rate at which
community composition shifts to favor species better able to take
advantage of, or cope with, new resource levels \citep{Smith2009}.
Previous work has shown that community compositional shifts can be
either rapid, on the order of years \citep{Hoover2014}, or slow, on the
order of decades \citep{Knapp2012, Wilcox2016}. A lingering question is
how the time scales of ecosystem response and community change vary
among ecosystems. Precipitation manipulation experiments can help answer
this question, especially if they push water availability outside the
historical range of variability for long periods.

@@

The similarity of ecosystem functional response (Fig. 3) and community
composition (Fig. 4) among treatments is surprising because grasslands
generally, and sagebrush steppe specifically, are considered
water-limited systems. \new{For example,} \citet{Huxman2004} and
\citet{Knapp2015}
\new{showed that semi-arid sites are more sensitive to drought than mesic sites, and}
\citet{Wilcox2017}
\new{found that semi-arid sites are particularly sensitive to irrigation treatments.}
\new{Based on these findings}, we expected ecosystem functional
response, community composition, or both to change under our treatments.
Why did our treatments fail to induce ecosystem or community responses?
We can think of three reasons. Two are limitations of our study, and one
invokes the life history traits of the species at our study site.

First, our precipitation manipulations may not have been large enough to
induce a response. A 50\% decrease in any given year may not be abnormal
given our site's historical range of variability \citep{Knapp2017}. We
cannot definitively rule out this possibility, but we have reason to
believe our treatments \emph{should} have been large enough. Using the
methods described by \citet{Lemoine2016}, we calculated the percent
reduction and increase of \new{water-year} precipitation necessary to
reach the 1\% and 99\% extremes of the historical precipitation regime
at our site (Fig. A5-1).
\new{The 1\% quantile of water-year precipitation at our site is 78 mm, a 26\% reduction from the mean, and the 99\% quantile is 545 mm, a 84\% increase from mean growing season precipitation (Appendix 6).
Thus, our drought treatment represented extreme precipitation amounts, and that is the treatment where we observed a small effect on the slope between soil moisture and ANPP (Fig. 3A).
The irrigation treatment may not have been extreme enough to induce a response.}

Second, ANPP at our site is likely influenced by additional factors, not
only the cumulative March-June soil moisture covariate we included in
our statistical model. \new{For example,} \citet{LaPierre2016}
\new{found that site-scale ANPP is better predicted by nutrient availability than precipitation.}
Moreover, temperature can impact ANPP directly \citep{Epstein1997} and
by exacerbating the effects of soil moisture \citep{DeBoeck2011}.
Measurements of soil moisture likely contain a signal of temperature,
through its impact on evaporation and infiltration, but the measurements
will not capture the direct effect of temperature on metabolic and
physiological processes. We also did not redistribute snow across our
treatments in the winter, and snow melt may spur early spring growth.
Failure to account for potentially important covariates could explain
the within-year spread of ANPP (Fig 2C, Fig. 3B) and the resulting
uncertain relationship we observed between soil moisture and ANPP across
all treatments (Table 1, Fig. A3-2).

Third, the life history traits of the dominant species in our study
ecosystem may explain the consistently positive, but weak and uncertain,
effect of soil moisture on ANPP (Table 1, Fig. 3). Species that live in
variable environments, such as cold deserts, must have strategies to
ensure long-term success as conditions vary. One strategy is bet
hedging, where species forego short-term gains to reduce the variance of
long-term success \citep{Seger1987}. In other words, species follow the
same conservative strategy every year, designed to minimize losses
during unfavorable periods. The dry and variable environment of the
sagebrush steppe has likely selected for bet hedging species that can
maintain function at low water availability and have weak responses to
high water availability. In so doing, the dominant species in our
ecosystem avoid ``boom and bust'' cycles, which may explain the weak
effect of soil moisture on ANPP (i.e., the Bayesian credible intervals
for the slopes overlapping zero; Table 1).

Another strategy to deal with variable environmental conditions is
avoidance, which would also result in a consistent ecosystem functional
response between drought and control treatments. For example, many of
the perennial grasses in our focal ecosystem avoid drought stress by
growing early in the growing season \citep[A.R. Kleinhesselink, personal
observation]{Blaisdell1958}. Furthermore, the dominant shrub in our
focal ecosystem, \emph{Artemisia tripartita}, has access to water deep
in the soil profile thanks to a deep root system
\citep{Kulmatiski2017a}.
\new{The dominance of our site by the shrub \emph{A. tripartita} (Appendix 1) may explain why our results do not conform to broader patterns of grassland sensitivity to precipitation manipulations}
\citep[e.g.,][]{Huxman2004, Knapp2015, Wilcox2017}.

@@

\section{CONCLUSIONS}\label{conclusions}

Our results suggest that the species in our focal plant community are
insensitive to to changes in precipitation regime,
\new{at least over the five years of our experiment}. Such insensitivity
can buffer species against precipitation variability in this semi-arid
ecosystem, making them successful in the long run. Longer, chronic
precipitation alteration might reveal plant community shifts that we did
not observe \citep[e.g.,][]{Wilcox2016},
\new{and the increased sensitivity of ANPP to drought in the final years of our experiment may portend such a shift.}
Likewise, a long-term increase in water availability could allow species
that do not bet hedge to gain prominence and dominate the ecosystem
functional response.
\new{The length of the perturbation may be especially relevant in our focal ecosystem where the perennial species are long-lived, meaning compositional turnover may take many more years than we report on here.}

\section{ACKNOWLEDGEMENTS}\label{acknowledgements}
